Analyzing changes in Canadian Grocery Prices

### 2024-11-21 paper/paper.qmd
Updated paper spacing

library(tidyverse)
library(ggplot2)
library(dplyr)
# library(knitr)
# library(rstanarm)
# library(arrow)
# library(data.table)
library(here)

@@

For this study, we utilize data from Project Hammer, a Canadian initiative designed to monitor grocery prices across major retailers in Canada. The Project Hammer dataset was collected from eight prominent Canadian grocery vendors—Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods—between February 28, 2024, and the latest available data load in the database [@hammer]. This dataset enables us to investigate price fluctuations, identify pricing trends, and explore potential competitive or collusive patterns within the Canadian grocery sector.
Project Hammer aims to support regulatory efforts by providing data on price transparency, making it suitable for academic research and potential legal analysis. Following data processing and cleaning procedures outlined in Section 3.2, the dataset was structured to allow an analysis of historical price changes and vendor-specific price patterns.
The dataset contains 1,996,969 rows and 5 columns, with variables capturing product (product name), vendor_name (name of the grocery retailer), price_current (current product price at the time of recording), price_old (previous recorded price for the product), and price_difference (difference between the current and old prices). Price data is recorded in Canadian dollars and captures a broad range of grocery items from various categories, including fresh produce, dairy, pantry staples, and household items.
To ensure the quality and consistency of the data, we focused on removing missing values and imputing outlier prices when necessary. For this analysis, we excluded entries with extreme price fluctuations beyond three standard deviations, as these could represent temporary discounts or data recording anomalies. This cleaning process provides a balanced and accurate reflection of grocery pricing, allowing for robust statistical analysis on vendor-specific trends and average price comparisons across product types.
By narrowing the scope to recent data, Project Hammer's dataset captures real-time fluctuations in grocery pricing, providing timely insights into potential market dynamics and consumer price impacts. The analysis allows for future investigation of temporal price trends, vendor-based price differences, and, more broadly, consumer affordability in a market facing rising cost-of-living pressures.

@@

\begin{align*}
    y_i \mid \mu_i, \sigma &\sim \text{Normal}(\mu_i, \sigma) \\
    \mu_i &= \beta_0 + \beta_1 \cdot \text{Previous Price}_i + \beta_2 \cdot \text{Vendor}_i + \beta_3 \cdot \text{Price Difference}_i \\
    \epsilon_i &\sim \text{Normal}(0, \sigma^2)
\end{align*}

\textbf{Where:}
\begin{itemize}
    \item $\beta_0$ is the intercept term, representing the baseline estimate of $price_{current}$
    \item $\beta_1$, $\beta_2$, and $\beta_3$ are the coefficients representing the effects of $price_{old}$, $vendor_{name}$ and $price_{difference}$ on $price_{current}$.
    \item $\sigma^2$represents the variance of the error term, capturing unexplained variability in current prices.
\end{itemize}

The model is executed in \texttt{R} [@citeR] using the \texttt{rstanarm} package [@rstanarm],  with priors set to regularize the estimates and prevent overfitting. Specifically, we use a normal prior for the coefficients, centered at zero with moderate variance, to ensure stable estimates without overly restrictive assumptions.

@@

## Interpretation of Grocery Pricing Dynamics
This study investigates the primary factors influencing current grocery prices in Canada by examining the Project Hammer dataset. Our Bayesian regression model reveals that prior prices (`price_old`), vendor identity (`vendor_name`), and recent price differences (`price_difference`) play significant roles in determining current prices. The influence of previous prices indicates price continuity across time, suggesting that grocery items tend to exhibit relatively stable pricing patterns with minor adjustments. Vendor-specific effects highlight the role of retailer pricing strategies, with some vendors maintaining consistently higher or lower prices. These vendor variations may reflect differences in market positioning, operational costs, or competitive strategies, impacting price consistency across the grocery sector.

>>>>

library(tidyverse)
library(ggplot2)
library(dplyr)
library(here)

@@

For this study, we utilize data from Project Hammer, a Canadian initiative designed to monitor grocery prices across major retailers in Canada. The Project Hammer dataset was collected from eight prominent Canadian grocery vendors—Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods—between February 28, 2024, and the latest available data can be found on the project hammer website [@hammer]. This dataset enables us to investigate price fluctuations, identify pricing trends, and explore potential competitive or collusive patterns within the Canadian grocery sector.

Project Hammer aims to support regulatory efforts by providing data on price transparency, making it suitable for academic research and potential legal analysis. Following data processing and cleaning procedures outlined in Section 3.2, the dataset was structured to allow an analysis of historical price changes and vendor-specific price patterns.

The dataset contains 1,996,969 rows and 5 columns, with variables capturing product (product name), vendor_name (name of the grocery retailer), price_current (current product price at the time of recording), price_old (previous recorded price for the product), and price_difference (difference between the current and old prices). Price data is recorded in Canadian dollars and captures a broad range of grocery items from various categories, including fresh produce, dairy, pantry staples, and household items.

To ensure the quality and consistency of the data, we focused on removing missing values and imputing outlier prices when necessary. For this analysis, we excluded entries with extreme price fluctuations beyond three standard deviations, as these could represent temporary discounts or data recording anomalies. This cleaning process provides a balanced and accurate reflection of grocery pricing, allowing for robust statistical analysis on vendor-specific trends and average price comparisons across product types.

By narrowing the scope to recent data, Project Hammer's dataset captures real-time fluctuations in grocery pricing, providing timely insights into potential market dynamics and consumer price impacts. The analysis allows for future investigation of temporal price trends, vendor-based price differences, and, more broadly, consumer affordability in a market facing rising cost-of-living pressures.

@@

```{=tex}
\begin{align*}
    y_i \mid \mu_i, \sigma &\sim \text{Normal}(\mu_i, \sigma) \\
    \mu_i &= \beta_0 + \beta_1 \cdot \text{Previous Price}_i + \beta_2 \cdot \text{Vendor}_i + \beta_3 \cdot \text{Price Difference}_i \\
    \epsilon_i &\sim \text{Normal}(0, \sigma^2)
\end{align*}
```
\textbf{Where:}

```{=tex}
\begin{itemize}
    \item $\beta_0$ is the intercept term, representing the baseline estimate of $price_{current}$
    \item $\beta_1$, $\beta_2$, and $\beta_3$ are the coefficients representing the effects of $price_{old}$, $vendor_{name}$ and $price_{difference}$ on $price_{current}$.
    \item $\sigma^2$represents the variance of the error term, capturing unexplained variability in current prices.
\end{itemize}
```
The model is executed in \texttt{R} [@citeR] using the \texttt{rstanarm} package [@rstanarm], with priors set to regularize the estimates and prevent overfitting. Specifically, we use a normal prior for the coefficients, centered at zero with moderate variance, to ensure stable estimates without overly restrictive assumptions.

@@

## Interpretation of Grocery Pricing Dynamics

This study investigates the primary factors influencing current grocery prices in Canada by examining the Project Hammer dataset. Our Bayesian regression model reveals that prior prices (`price_old`), vendor identity (`vendor_name`), and recent price differences (`price_difference`) play significant roles in determining current prices. The influence of previous prices indicates price continuity across time, suggesting that grocery items tend to exhibit relatively stable pricing patterns with minor adjustments. Vendor-specific effects highlight the role of retailer pricing strategies, with some vendors maintaining consistently higher or lower prices. These vendor variations may reflect differences in market positioning, operational costs, or competitive strategies, impacting price consistency across the grocery sector.
